### 2025-04-21 evaluating-prediction-accuracy.qmd
20250421 - example of prediction of categorical outcome

>>>>

## Example of Predicting a Categorical outcome {#sec-predictionCategoricalExample}

Let's consider an example of predicting a categorical outcome.
For this example, let's predict whether or not a Running Back rushes for 200 yards in a game.
To predict that, we will use the number of rushing yards the player had in the previous game.

```{r}
player_stats_weekly_example <- player_stats_weekly %>% 
  filter(position == "RB") %>% 
  arrange(player_id, season, week) %>% 
  group_by(player_id, season) %>% 
  mutate(rushing_yards_priorGame = lag(rushing_yards)) %>% 
  ungroup()
```

```{r}
player_stats_weekly_example %>% 
  select(player_id, season, week, rushing_yards, rushing_yards_priorGame) %>% 
  head()
```

Here are our accuracy indices at a cutoff of 100.
That is, we predict that each player who rushed more than 100 yards in a given game will rush for 200 yards in their next game.

```{r}
petersenlab::accuracyAtCutoff(
  predicted = player_stats_weekly_example$rushing_yards_priorGame,
  actual = player_stats_weekly_example$rush_200_yds,
  cutoff = 100
)
```

Now, let's lower the cutoff to 50 and see how that impacts our predictive accuracy.

```{r}
petersenlab::accuracyAtCutoff(
  predicted = player_stats_weekly_example$rushing_yards_priorGame,
  actual = player_stats_weekly_example$rush_200_yds,
  cutoff = 50
)
```

Now, let's raise the cutoff to 150 and see how that impacts our predictive accuracy.

```{r}
petersenlab::accuracyAtCutoff(
  predicted = player_stats_weekly_example$rushing_yards_priorGame,
  actual = player_stats_weekly_example$rush_200_yds,
  cutoff = 150
)
```

We can also examine the accuracy at each possible cutoff.

```{r}
petersenlab::accuracyAtEachCutoff(
  predicted = player_stats_weekly_example$rushing_yards_priorGame,
  actual = player_stats_weekly_example$rush_200_yds
)
```

We can also create a receiver operating characteristic (ROC) curve, as in @fig-rocCurveExample.

```{r}
rocCurve <- pROC::roc(
  data = player_stats_weekly_example,
  response = rush_200_yds,
  predictor = rushing_yards_priorGame,
  smooth = FALSE)
```

Here is the area under the ROC curve (AUC):

```{r}
rocCurve$auc
```

```{r}
#| label: fig-rocCurveExample
#| fig-cap: "Example Receiver Operating Characteristic (ROC) Curve."
#| fig-alt: "Example Receiver Operating Characteristic (ROC) Curve."
#| fig-width: 8
#| fig-height: 6

plot(
  rocCurve,
  legacy.axes = TRUE,
  print.auc = TRUE)
```
